---
title: "ENV 790.30 - Time Series Analysis for Energy Data | Spring 2025"
subtitle: "Assignment 2 - Due date 01/23/25"
author: "Justin Maynard"
output: pdf_document
geometry: margin=2.54cm
editor_options: 
  chunk_output_type: console
---

## Submission Instructions

You should open the .rmd file corresponding to this assignment on RStudio. The file is available on our class repository on Github.   

Once you have the file open on your local machine the first thing you will do is rename the file such that it includes your first and last name (e.g., "LuanaLima_TSA_A02_Sp24.Rmd"). Then change "Student Name" on line 4 with your name. 

Then you will start working through the assignment by  **creating code and output** that answer each question. Be sure to use this assignment document. Your report should contain the answer to each question and any plots/tables you obtained (when applicable).

When you have completed the assignment, **Knit** the text and code into a single PDF file. Submit this pdf using Sakai. 

## R packages

R packages needed for this assignment:"forecast","tseries", and "dplyr". Install these packages, if you haven't done yet. Do not forget to load them before running your script, since they are NOT default packages.\\

```{r message=FALSE, warning=FALSE}
#Load/install required package here
library(forecast)
library(tseries)
library(dplyr)
library(here)
library(readxl)
library(ggplot2)
```


## Data set information  

Consider the data provided in the spreadsheet "Table_10.1_Renewable_Energy_Production_and_Consumption_by_Source.xlsx" on our **Data** folder. The data comes from the US Energy Information and Administration and corresponds to the December 2023 Monthly Energy Review. The spreadsheet is ready to be used. You will also find a $.csv$ version of the data "Table_10.1_Renewable_Energy_Production_and_Consumption_by_Source-Edit.csv". You may use the function $read.table()$ to import the $.csv$ data in R. Or refer to the file "M2_ImportingData_CSV_XLSX.Rmd" in our Lessons folder for functions that are better suited for importing the $.xlsx$.

```{r message=FALSE, warning=FALSE}
#Importing data set
data <- read_excel("Data/Table_10.1_Renewable_Energy_Production_and_Consumption_by_Source.xlsx", 
                   skip = 12, col_names = FALSE)
names <- read_excel("Data/Table_10.1_Renewable_Energy_Production_and_Consumption_by_Source.xlsx",
                    skip = 10, n_max = 1, col_names = F)
colnames(data) <- names

```


## Question 1
You will work only with the following columns: Total Biomass Energy Production, Total Renewable Energy Production, Hydroelectric Power Consumption. Create a data frame structure with these three time series only. Use the command head() to verify your data.

```{r}
data <- data[,c("Month", "Total Biomass Energy Production", 
                "Total Renewable Energy Production", "Hydroelectric Power Consumption")]
head(data)

```


## Question 2
Transform your data frame in a time series object and specify the starting point and frequency of the time series using the function ts().

```{r}
ts <- ts(data[,2:4], start = c(1973,1,1), frequency = 12)
head(ts)
```


## Question 3
Compute mean and standard deviation for these three series.

```{r}

print(paste("Mean of",colnames(ts)[1], "is",round(mean(ts[,1]),2)))
print(paste("Mean of",colnames(ts)[2], "is",round(mean(ts[,2]),2)))
print(paste("Mean of",colnames(ts)[3], "is",round(mean(ts[,3]),2)))


print(paste("Standard deviation of",colnames(ts)[1], "is",round(sd(ts[,1]),2)))
print(paste("Standard deviation of",colnames(ts)[2], "is",round(sd(ts[,2]),2)))
print(paste("Standard deviation of",colnames(ts)[3], "is",round(sd(ts[,3]),2)))

```


## Question 4
Display and interpret the time series plot for each of these variables. Try to make your plot as informative as possible by writing titles, labels, etc. For each plot add a horizontal line at the mean of each series in a different color.

Based on the following plot, hydroelectric power consumption varies mostly seasonally, and stays near it's average with little trend. Biomass energy production and renewable energy production also vary seasonally, but have an increasing trend.

```{r}

ggplot(data, aes(x = Month)) + 
  geom_line(aes(y = `Total Biomass Energy Production`, col = "Biomass")) +
  geom_hline(yintercept = mean(data$`Total Biomass Energy Production`), color = 'green') +
  geom_line(aes(y = `Total Renewable Energy Production`, col = "Renewable")) +
  geom_hline(yintercept = mean(data$`Total Renewable Energy Production`), color = 'blue') +
  geom_line(aes(y = `Hydroelectric Power Consumption`, col = "Hydroelectric")) + 
  geom_hline(yintercept = mean(data$`Hydroelectric Power Consumption`), color = 'red') +
  xlab("Date") + 
  ylab("Renewable Energy Production (Trillion Btu)") + 
  labs(color = "Energy Type") + 
  theme(legend.position = "bottom") 
  

```


## Question 5
Compute the correlation between these three series. Are they significantly correlated? Explain your answer.

Renewable energy production is highly correlated to biomass energy production (0.97), which may be due to the similar increasing trend that biomass and renewable production follow, as well as similar seasonal cycles. Hydroelectric power consumption is not strongly correlated with biomass production (-0.11) or renewable energy production (-0.03) as hydroelectric power consumption does not show the same increasing trend, as new hydroelectric plants are rarely built.

```{r}
print(cor(ts))
```


## Question 6
Compute the autocorrelation function from lag 1 up to lag 40 for these three variables. What can you say about these plots? Do the three of them have the same behavior?

The autocorrelation plots for biomass and renewable energy have the same behavior of slight decrease in autocorrelation from lag 1 to lag 40. Hydroelectric power production follows a different trend, and oscillates between positive and negative autocorrelation following a wave shape. This is because there is little trend and high seasonality. 

```{r}
#Biomass
Acf_1 <- Acf(ts[,1],lag=40)
#Renewable
Acf_2 <- Acf(ts[,2],lag=40)
#Hydro
Acf_3 <- Acf(ts[,3],lag=40)

```


## Question 7
Compute the partial autocorrelation function from lag 1 to lag 40 for these three variables. How these plots differ from the ones in Q6?

These plots differ in that they all start with a high PACF and then exhibit the "wave" pattern found in the hydroelectric plot in A6, except that the wave pattern is decreasing in magnitude as lag increases.

```{r}
Pacf_1 <- Pacf(ts[,1],lag=40)
Pacf_2 <- Pacf(ts[,2],lag=40)
Pacf_3 <- Pacf(ts[,3],lag=40)

```


